Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-12-11 Manuscript/manuscriptPSsuppl.tex
Update manuscript

\noindent
{\it POPC bilayer}
\todo{By Fernando Favela at http://dx.doi.org/10.5281/zenodo.16830} \\
% It should be noted that pure POPC are ran with Gromacs 4, but the results with
% PS are ran with Gromacs 5. This has significant effect on acyl chains:
% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
% The effect for headgroups can be estimated by comparing these two data sets:
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}

@@

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of other lipids with bound charge,
like negatively charged PS and PG lipids,  are also systematic,
but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

When comparing ion binding measured from the PC headgroup order parameters between
neutral and anionic membranes, it is important to note that the order parameters
increase due to the addition of negative charged lipids according to the electrometer
concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
Therefore, the headgroup order parameters of POPC in mixtures with anionic
lipids are larger than in pure POPC lipid bilayer, which is also seen in the ion binding
experiments from different mixtures in the absence of added calcium (Fig. \ref{OrderParametersWithCaCl}).
Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
the negative charge lipids, resulting to a neutral membrane. 
Above such concentrations, the specific binding of calcium leads to
the overcharging of the membrane.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    Headgroup order parameters of POPC as a function of CaCl concentration from experiments 
    different mole fractions of negatively charged lipids (left column).
    Right column shows the same data zoomed to the concentrations below 1M.
    Data for Pure DPPC from~\cite{akutsu81},
    for pure POPC from~\cite{altenbach84}, 
    for POPC:POPS (5:1) mixture from~\cite{roux90},
    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from~\cite{macdonald87}
    and for POPC:POPG (1:1) mixture data without NaCl from \cite{borle85}.
  }
\end{figure}

@@

The dihedral angles and structures of the glycerol backbone and headgroup regions of POPS lipids show
wide variety between different simulation models (Figs. \ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
Detailed discussion of the structural differences is limited by the lack of realistic model that would correctly reproduce the
headgroup and glycerol backbone order parameters. However, some structural characteristics
of PS headgroup can be suggested based on the best available models
(Figs. \ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
The glycerol backbone structures are not discussed in this works because our focus is on PS headgroup.
However, the data presented here can be useful for future investigations.
%backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
%differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
%as observed previously also for PC lipids \cite{botan15}.
%The origin of this difference is more difficult to track without more elaborate analysis,
%because different models show very complicated patterns of distinct structures
%in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).

>>>>

\noindent
    {\it POPC bilayer.} Previously published values \cite{botan15} calculated from the data
    available from Ref. \citenum{charmmPOPC512files} are used.
% It should be noted that pure POPC are ran with Gromacs 4, but the results with
% PS are ran with Gromacs 5. This has significant effect on acyl chains:
% https://github.com/NMRLipids/NmrLipidsCholXray/issues/4
% but the effect to headgroups is small
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS4.5/Headgroup_Glycerol_Order_Parameters_Simulation.dat
% https://github.com/NMRLipids/MATCH/blob/master/Data/Lipid_Bilayers/POPC/T303K/MODEL_CHARMM36_GROMACS5.1.2/Headgroup_Glycerol_Order_Parameters_Simulation.dat}

@@

The electrometer concept is based on the empirical observations that
the order parameters of $\alpha$ and $\beta$ carbons in PC lipid headgroup
decrease (increase) proportionally to the bound positive (negative) 
charge~\cite{akutsu81,altenbach84,seelig87,scherer89} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}). 
Therefore, the headgroup order parameters can be used to measure the
ion binding affinity to lipid bilayers containing PC
lipids~\cite{akutsu81,altenbach84,borle85,seelig87,macdonald87,roux90}.
Changes of the headgroup order parameters of negatively charged PS and PG lipids
are also systematic, but less well characterized~\cite{borle85,macdonald87,roux86,roux90}.
Therefore, the ion binding affinity to negatively charged bilayers
can be better characterized measuring the PC headgroup order parameters from 
mixed bilayers~\cite{borle85,roux86,macdonald87,roux90,roux91}.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPEPSPGchol.eps}
  \caption{\label{HGorderparametersPCvsPEPSPGchol}
    Headgroup order parameters of POPC measured from mixtures with
    PS (bovine brain), POPG, POPE, cholesterol and cationic dihexadecyldimethylammonium bromide (DHAB) surfactant~\cite{scherer87,scherer89,ferreira13}.
    Signs are taken from separate experiments~\cite{ollila16,ferreira16}.
  }
\end{figure}

When using the PC headgroup order parameters to evaluate the ion binding affinity to a bilayer containing anionic lipids,
it is important to note that the order parameters
increase due to the addition of negative charged lipids according to the electrometer
concept~\cite{seelig87, scherer87} (Fig. \ref{HGorderparametersPCvsPEPSPGchol}).
Therefore, the PC headgroup order parameters are larger in mixtures with anionic lipids
than in pure lipid bilayers. This is evident also in the headgroup order parameters
of POPC in bilayers with different amounts anionic lipids without added calcium (Fig. \ref{OrderParametersWithCaCl}).
Upon addition of CaCl$_2$, the order parameters decrease and reach the values of pure PC bilayer close 
to the CaCl$_2$ concentrations of $\sim$ 50-300mM, depending on the amount of negatively charged
lipids in the mixture (Fig. \ref{OrderParametersWithCaCl}).
Around these concentrations the positive charge of bound Ca$^{2+}$ cancels
the negative charge lipids, resulting to a neutral membrane. 
Above such concentrations, the specific binding of calcium leads to
the overcharging of the membrane.
\begin{figure}[]
  \centering
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaCl.eps}
  \includegraphics[width=8.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
  \caption{\label{OrderParametersWithCaCl}
    Headgroup order parameters of POPC as a function of CaCl$_2$ concentration from experiments 
    with different mole fractions of negatively charged lipids (left column).
    Right column shows the same data zoomed to the concentrations below 1M.
    Data for Pure DPPC from Ref.~\citenum{akutsu81},
    for pure POPC from Ref.~\citenum{altenbach84}, 
    for POPC:POPS (5:1) mixture from Ref.~\citenum{roux90},
    for POPC:POPG (8:2,1:1) mixtures with 0.1M NaCl from Ref.~\citenum{macdonald87}
    and for POPC:POPG (1:1) mixture data without NaCl from Ref.\citenum{borle85}.
  }
\end{figure}

@@

The dihedral angles and structures of the glycerol backbone and headgroup regions of POPS lipids show
wide variety between different simulation models (Figs. \ref{dihedralsHG}, \ref{dihedralsGLY} and \ref{HGandGLYstructuresPS}).
Detailed discussion of the structural differences is limited by the lack of realistic model that would correctly reproduce the
headgroup and glycerol backbone order parameters. However, some structural characteristics
of PS headgroup can be suggested based on the best available models
(Figs. \ref{dihedralsHGpc}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}), as discussed in the main text.
The glycerol backbone structures are not discussed in this work because our focus is on PS headgroup.
However, the data presented here can be useful for future investigations.
%backbone order parameters of C$_2$ and C$_3$ from Slipids simulations
%differ significantly from the other simulation results and experiments (Fig. \ref{HGorderParametersPS}),
%as observed previously also for PC lipids \cite{botan15}.
%The origin of this difference is more difficult to track without more elaborate analysis,
%because different models show very complicated patterns of distinct structures
%in the glycerol backbone region (Figs.~\ref{HGstructuresPS} and \ref{dihedralsHG}).
